Perturbing SU($n$)-symmetric interactions

### 2020-02-10 collective_sun/sun_perturbation.tex
simplify the two-body problem in the case of an isotropic system

\newcommand{\PP}{\mathbb{P}}
\renewcommand{\SS}{\mathbb{S}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\ZZ}{\mathbb{Z}}

@@

Here we simplify the eigenvalue problems that appear at second order
in the single- and two-body theory in Sections
\ref{sec:single_body_pert} and \ref{sec:two_body_pert} in the case of
a translationally invariant system with periodic boundary conditions.
We solve the single-body eigenvalue problem analytically, and reduce
the two-body eigenvalue problem to that of diagonalizing
$\p{N-1}\times\p{N-1}$ matrices.  Finally, for we discuss the further
reduction of the two-body problem to that of diagonalizing matrices
with dimensions $\sim\p{N/2^D}\times\p{N/2^D}$ in the case of an
isotropic system on a $D$-dimensional periodic lattice, but leave this
reduction to future work.

@@

The two-body eigenvalue equation in \eqref{eq:cond_2} is designed to
find solutions $\m w$ to the equation
\begin{align}
  \sum_k \p{h_{pk} w_{kq} + w_{pk} h_{kq}}
  - \bar h_{pq} w_{pq}
  = \Delta w_{pq},
  &&
  \bar h_{pq} \equiv h_p + h_q - 2h_{pq},
  &&
  h_p \equiv \sum_q h_{pq},
  \label{eq:cond_2_ref_start}
\end{align}
which can be written in matrix form as
\begin{align}
  \P_{\t{off-diag}}\p{\m h\c\m w + \m w\c\m h} - \bar{\m h} * \m w
  = \Delta \m w,
  \label{eq:cond_2_ref}
\end{align}
where $\m h$, $\bar{\m h}$, and $\m w$ are matrices of all $h_{pq}$,
$\bar h_{pq}$ and $w_{pq}$;
$\P_{\t{off-diag}}\p{\m m}\equiv \m m-\sum_p\op{p}\m m\op{p}$ denotes
the projection of $\m m$ onto the space of strictly off-diagonal
matrices; and $\bar{\m h}*\m w$ denotes an element-wise (Kronecker)
product of $\bar{\m h}$ and $\m w$, i.e.~with matrix elements
$\bar h_{pq}w_{pq}$.  In order to find solutions to
\eqref{eq:cond_2_ref} when the interaction Hamiltonian $H_0$ is
translationally invariant, in which case $h_{pq}$ and $\bar h_{pq}$
depend only on the separation $\pm\p{p-q}$, we expand all matrices in
a basis that the respects translational symmetry.  To this end, for
all lattice displacements $d\in\ZZ_L^D$ and wavenumbers
$k\in\ZZ_L^D\times2\pi/L$ we define the {\it cycle matrices}
\begin{align}
  \m m_{dk} \equiv \sum_p e^{ip\c k} \op{p+d}{p},
  \label{eq:cycle}
\end{align}
where the element-wise sums in $p+d$ are implicitly taken modulo $L$
(i.e.~the length of the periodic lattice along each axis).  In terms
of these cycle matrices, we define
\begin{align}
  \tilde w_{dk} \equiv \f1N \tr\p{\m m_{dk}^\dag \m w},
\end{align}
and expand
\begin{align}
  \m w = \sum_{d,k} \tilde w_{dk} \m m_{dk},
  &&
  \m h = \sum_{q,d} h_{q+d,q} \op{q+d}{q}
  = \sum_d h_{d,0} \m m_{d,0},
  &&
  \bar{\m h} = \sum_d \bar h_{d,0} \m m_{d,0}.
\end{align}
We can then write the condition in \eqref{eq:cond_2_ref} as a new set
of conditions
\begin{align}
  \sum_{\substack{c,d\\c+d\ne0}}
  h_{c,0} \tilde w_{dk} \p{1 + e^{-ic\c k}} \m m_{c+d,k}
  - \sum_{d\ne0} \bar h_{d,0} \tilde w_{dk} \m m_{dk}
  = \Delta \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
\end{align}
that must hold for all wavenumbers $k$.  With some re-indexing, we can
write
\begin{align}
  \sum_{c,d\ne0} h_{d-c,0} \tilde w_{ck}
  \p{1 + e^{-i\p{d-c}\c k}} \m m_{dk}
  - \sum_{d\ne0} \bar h_{d,0} \tilde w_{dk} \m m_{dk}
  = \Delta \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
  \label{eq:cond_2_cycle}
\end{align}
and define
\begin{align}
  \ut{\check{\m h}}_k
  \equiv \sum_{c,d\ne0} h_{d-c,0} \p{1+e^{-i\p{d-c}\c k}} \op{d}{c}
  - \sum_{d\ne0} \bar h_{d,0} \op{d},
  &&
  \ut{\v{\m w}}_k \equiv \sum_{d\ne0} \tilde w_{dk} \ket{d},
  \label{eq:cycle_vecs}
\end{align}
in order to express \eqref{eq:cond_2_cycle} as the set of eigenvalue
equations
\begin{align}
  \ut{\check{\m h}}_k \c \ut{\v{\m w}}_k = \Delta \ut{\v{\m w}}_k.
  \label{eq:cond_2_square}
\end{align}
The appearance of a wavenumber index $k$ in \eqref{eq:cond_2_square}
implies that all fixed-$k$ components of a general coefficient matrix
$\m w$ must simultaneously satisfy \eqref{eq:cond_2_square} in order
to generate states of fixed excitation energy $\Delta$ when the
corresponding perturbation $\V_2^\Delta$ acts on a permutationally
symmetric state $\ket\psi\in\M_0$.  The decomposition of an arbitrary
coefficient vector $\v{\m w}=\sum_\Delta\v{\m w}_\Delta$ into
coefficient vectors $\v{\m w}_\Delta$ that generate a
fixed-excitation-energy perturbation $\V_2^\Delta$ is then performed
by projecting each $k$-component of $\v{\m w}$ onto the corresponding
$\Delta$-eigenspace of $\ut{\check{\m h}}_k$.  If the perturbation
$\V_2$ is translationally invariant, then the $k$-components of
$\v{\m w}$ for all $k\ne 0$ all vanish, so we only need to consider
the case of $k=0$.

We can reduce the eigenvalue problem in \eqref{eq:cond_2_square} even
further in the case of an isotropic system, in which the couplings
$h_{pq}$ depend only on the distance $\abs{p-q}$.  On a $D$
dimensional lattice, isotropy implies one reflection symmetry for each
lattice axis, which reduces the degrees of freedom in of
$\ut{\check{\m h}}_k$ from $\sim N\times N$ to
$\sim\p{N/2^D}\times\p{N/2^D}$.  We leave this reduction to future
work.

>>>>

\newcommand{\DD}{\mathbb{D}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\renewcommand{\SS}{\mathbb{S}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\ZZ}{\mathbb{Z}}

@@

Here we simplify the eigenvalue problems that appear at second order
in the single- and two-body theory in Sections
\ref{sec:single_body_pert} and \ref{sec:two_body_pert} in the case of
a translationally invariant system with periodic boundary conditions.
We solve the single-body eigenvalue problem analytically, and reduce
the two-body eigenvalue problem to that of diagonalizing
$\p{N-1}\times\p{N-1}$ matrices.  We then reduce the two-body
eigenvalue problem even further for the case of an isotropic system,
to that of diagonalizing a single $\sim\p{N/2^D}\times\p{N/2^D}$
matrix.  Finally, we provide an analytic solution to the two-body
problem in the case of a fully collective model with uniform
couplings.

@@

The two-body eigenvalue equation in \eqref{eq:cond_2} is designed to
find solutions $\m w$ to the equation
\begin{align}
  \sum_k \p{h_{pk} w_{kq} + w_{pk} h_{kq}} - \bar h_{pq} w_{pq}
  = \Delta w_{pq},
  &&
  \bar h_{pq} \equiv h_p + h_q - 2h_{pq},
  &&
  h_p \equiv \sum_q h_{pq},
  \label{eq:cond_2_ref_start}
\end{align}
where translational invariance implies that
$\bar h_{pq}=2h_0-2h_{pq}$, so we can equivalently write
\begin{align}
  h_{pq} w_{pq}
  + \f12 \sum_k \p{h_{pk} w_{kq} + w_{pk} h_{kq}}
  = \p{\f12 \Delta + h_0} w_{pq},
\end{align}
which can also be written in matrix form as
\begin{align}
  \m h * \m w + \f12 \P_{\t{off-diag}}\p{\m h\c\m w + \m w\c\m h}
  = \p{\f12 \Delta + h_0} \m w,
  \label{eq:cond_2_ref}
\end{align}
where $\m h$ and $\m w$ are matrices of all $h_{pq}$ and $w_{pq}$;
$\m h*\m w$ denotes an element-wise (Kronecker) product of $\m h$ and
$\m w$, i.e.~with matrix elements $h_{pq}w_{pq}$; and
$\P_{\t{off-diag}}\p{\m m}\equiv \m m-\sum_p\op{p}\m m\op{p}$ denotes
the projection of $\m m$ onto the space of strictly off-diagonal
matrices.  In order to find solutions to \eqref{eq:cond_2_ref} when
the interaction Hamiltonian $H_0$ is translationally invariant, in
which case $h_{pq}$ depends only on the separation $\pm\p{p-q}$, we
expand all matrices in a basis that the respects translational
symmetry.  To this end, for all lattice displacements $d\in\ZZ_L^D$
and wavenumbers $k\in\ZZ_L^D\times2\pi/L$ we define the {\it cycle
  matrices}
\begin{align}
  \m m_{dk} \equiv \sum_p e^{ip\c k} \op{p+d}{p},
  \label{eq:cycle}
\end{align}
where the element-wise sums in $p+d$ are implicitly taken modulo $L$
(i.e.~the length of the periodic lattice along each axis).  In terms
of these cycle matrices, we define
\begin{align}
  \tilde w_{dk} \equiv \f1N \tr\p{\m m_{dk}^\dag \m w},
\end{align}
and expand
\begin{align}
  \m w = \sum_{d,k} \tilde w_{dk} \m m_{dk},
  &&
  \m h = \sum_{q,d} h_{q+d,q} \op{q+d}{q}
  = \sum_d h_{d,0} \m m_{d,0}.
\end{align}
We can then write the condition in \eqref{eq:cond_2_ref} as a new set
of conditions
\begin{align}
  \sum_{d\ne0} h_{d,0} \tilde w_{dk} \m m_{dk}
  + \f12 \sum_{\substack{c,d\\c+d\ne0}}
  h_{c,0} \tilde w_{dk} \p{1 + e^{-ic\c k}} \m m_{c+d,k}
  = \p{\f12\Delta + h_0} \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
\end{align}
that must hold for all wavenumbers $k$.  With some re-indexing, we can
write
\begin{align}
  \sum_{d\ne0} h_{d,0} \tilde w_{dk} \m m_{dk}
  + \f12 \sum_{c,d\ne0} h_{d-c,0} \tilde w_{ck}
  \p{1 + e^{-i\p{d-c}\c k}} \m m_{dk}
  = \p{\f12\Delta + h_0} \sum_{d\ne0} \tilde w_{dk} \m m_{dk},
  \label{eq:cond_2_cycle}
\end{align}
and define
\begin{align}
  \ut{\check{\m h}}_k
  \equiv \sum_{d\ne0} h_{d,0} \op{d}
  + \f12 \sum_{c,d\ne0} h_{d-c,0} \p{1+e^{-i\p{d-c}\c k}} \op{d}{c},
  &&
  \ut{\v{\m w}}_k \equiv \sum_{d\ne0} \tilde w_{dk} \ket{d},
  \label{eq:cycle_vecs}
\end{align}
in order to express \eqref{eq:cond_2_cycle} as the set of eigenvalue
equations
\begin{align}
  \ut{\check{\m h}}_k \c \ut{\v{\m w}}_k
  = \p{\f12\Delta + h_0} \ut{\v{\m w}}_k.
  \label{eq:cond_2_cycle_block}
\end{align}
The appearance of a wavenumber index $k$ in
\eqref{eq:cond_2_cycle_block} implies that all fixed-$k$ components of
a general coefficient matrix $\m w$ must simultaneously satisfy
\eqref{eq:cond_2_cycle_block} in order to generate states of fixed
excitation energy $\Delta$ when the corresponding perturbation
$\V_2^\Delta$ acts on a permutationally symmetric state
$\ket\psi\in\M_0$.  The decomposition of an arbitrary coefficient
vector $\v{\m w}=\sum_\Delta\v{\m w}_\Delta$ into coefficient vectors
$\v{\m w}_\Delta$ that generate a fixed-excitation-energy perturbation
$\V_2^\Delta$ is then performed by projecting each $k$-component of
$\v{\m w}$ onto the corresponding $\Delta$-eigenspace of
$\ut{\check{\m h}}_k$.  If the perturbation $\V_2$ is translationally
invariant, then $\ut{\v{\m w}}_k=0$ for all $k\ne0$, so we only need
to consider the case of $k=0$:
\begin{align}
  \ut{\check{\m h}}_0 \c \ut{\v{\m w}}_0
  = \p{\f12\Delta + h_0} \ut{\v{\m w}}_0,
  &&
  \ut{\check{\m h}}_0
  = \sum_{d\ne0} h_{d,0} \op{d}
  + \sum_{c,d\ne0} h_{cd} \op{c}{d},
  &&
  \ut{\v{\m w}}_0 \equiv \sum_{d\ne0} \tilde w_{d,0} \ket{d}.
  \label{eq:cond_2_cycle_0}
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Isotropic systems}

If the interaction Hamiltonian $H_0$ and two-body perturbation $\V_2$
are both translationally invariant isotropic, then we can reduce the
eigenvalue problem \eqref{eq:cond_2_cycle_0} even further by making
use of the fact that the couplings $h_{pq}$ depend only on the
distance $\abs{p-q}$.  To this end, we define vectors $\ket{a}$
labeled by distances $a\in\RR$ from the origin:
\begin{align}
  \ket{a} \equiv \f1{\sqrt{N_a}}
  \sum_{\substack{d\in\ZZ_L^D\\\abs{d}=a}} \ket{d},
\end{align}
where the norm $\abs{d}$ of a vector $d$ on a periodic lattice is
implicitly understood to mean the smallest Euclidean distance of $d$
from the origin; and $N_a$ is a normalization factor equal to the
number of displacements $d\in\ZZ_L^D$ with $\abs{d}=a$, such that
$\bk{a|a}=1$.  Defining the set of distances
\begin{align}
  \DD_{LD} \equiv \set{ \abs{d} : d \in \ZZ_L^D } \setminus \set{0},
\end{align}
we can expand
\begin{align}
  \ut{\check{\m h}}_0
  = \sum_{a\in\DD_{LD}} \tilde h_a \op{a}
  + \sum_{a,b\in\DD_{LD}} \tilde h_{ab} \op{a}{b},
  &&
  \ut{\v{\m w}}_0
  \equiv \sum_{a\in\DD_{LD}} \tilde w_a \sqrt{N_a} \ket{a},
\end{align}
where $\tilde h_a=h_{d,0}$ and
$\tilde w_a=\tilde w_{d,0}=\tr\p{\m m_{d,0}^\dag \m w}/N$ for all
$\abs{d}=a$, and
\begin{align}
  \tilde h_{ab} \equiv \f1{\sqrt{N_aN_b}}
  \sum_{\substack{c,d\in\ZZ_L^D\\\abs{c}=a\\\abs{d}=b}} h_{cd}.
\end{align}
By rough accounting, isotropy in $D$ dimensions implies one reflection
symmetry for each of $D$ axes, so the matrix $\ut{\check{\m h}}_0$ has
dimensions $\sim\p{N/2^D}\times\p{N/2^D}$.
